Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-06 paper/main.tex
Updates

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations} 
\todo[inline]{Needs improvement}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association. 
where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, $ \mathrm{corr}(X, Y) $ is the correlation between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively.  Covariance matrix using $ \Sigma $ and the entry corresponding to
covariance between $ X $ and $ Y $ in the matrix using $ \Sigma_{XY} $.

@@

We use Pillai's Trace because: 1) It uses all the canonical correlation values,
capturing the full extent of association, 2) Its interpretation is similar to
Pearson's correlation coefficient, i.e., $ 0 $ signifies no association and $ 1
$ signifies perfect linear relationship. We use a normalized version because
when comparing between pair of variables with different number of categories.

Our proposed measure has several desirable properties that make it well-suited
for our application:

\begin{enumerate}
	\item \textbf{Bounded: } The measure is bounded between $ 0 $ (no
		association) to $ 1 $ (perfect linear association), simplifying
		interpretation.
	\item \textbf{Independent of Sample Size: } 
	\item \textbf{Invariant to the Dummy Encoding: } For categorical
		variables, residuals are computed using dummy encoding. Since
		canonical correlations identifies linear combinations across
		columns, this measure is invariant to the specific dummy
		encoding scheme used.
	\item \textbf{Equivalent to Partial Correlation for Continuous Variables: }
		When both $ X $ and $ Y $ are continuous variables, our measure
		is equivalent to the absolute value of partial correlation coefficient.
	\item \textbf{Equivalent to polychoric and polyserial correlation for
			ordinal and continuous variables: }
		Under the assumption that the ordinal variable is generated by
		discretizing an underlying gaussian continuous variable, this
		effect size is equivalent to polychoric and polyserial
		correlation. As both of them recover the Pearson correlation
		coefficient.
	\item \textbf{Related to Cram\'er's V: } 
\end{enumerate}

>>>>

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations} 
\todo[inline]{Need to rewrite this}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association. 
where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, $ \mathrm{corr}(X, Y) $ is the correlation between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively.  Covariance matrix using $ \Sigma $ and the entry corresponding to
covariance between $ X $ and $ Y $ in the matrix using $ \Sigma_{XY} $.

@@

We use Pillai's Trace because: 1) It uses all the canonical correlation values,
capturing the full extent of association, 2) Its interpretation is similar to
Pearson's correlation coefficient, i.e., $ 0 $ signifies no association and $ 1
$ signifies perfect linear relationship. We use a normalized version because
when comparing between pair of variables with different number of categories.
Our proposed measure has several desirable properties that make it well-suited
for our application:

\begin{enumerate}
	\item \textbf{Bounded: } The measure is bounded between $ 0 $ (no
		association) to $ 1 $ (perfect linear association), simplifying
		interpretation.
	% \item \textbf{Independent of Sample Size: } 
	\item \textbf{Invariant to the Dummy Encoding: } For categorical
		variables, residuals are computed using dummy encoding. Since
		canonical correlations identifies linear combinations across
		columns, this measure is invariant to the specific dummy
		encoding scheme used.
	\item \textbf{Equivalent to Partial Correlation for Continuous Variables: }
		When both $ X $ and $ Y $ are continuous variables, our measure
		is equivalent to the absolute value of partial correlation coefficient.
	\item \textbf{Equivalent to polychoric and polyserial correlation for
			ordinal and continuous variables: }
		Under the assumption that the ordinal variable is generated by
		discretizing an underlying gaussian continuous variable, this
		effect size is equivalent to polychoric and polyserial
		correlation. As both of them recover the Pearson correlation
		coefficient.
	\item \textbf{Related to Cram\'er's V: } 
\end{enumerate}
